Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-08-14 ms/ms.qmd
addressing k8's comments

---
title: Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves
format:
  pdf:
    keep-tex: true  
    toc: false
author:
  - name: "Christopher D. Muir$^{1,2*}$"
    orcid: 0000-0003-2555-3878
  - name: "Wei Shen Lim$^{1}$"
  - name: "Dachuan Wang$^{1,2}$"
affiliations: false
number-sections: false
header-includes: |
  \usepackage{hyperref}
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
  \newcommand{\aalow}{\mathrm{AA}_{150}}
  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
  \newcommand{\aax}{$\mathrm{AA}$}
  \newcommand{\aay}{\mathrm{AA}}
  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
  \newcommand{\aqcurve}{$A \textendash Q$}
  \newcommand{\ca}{$C_\mathrm{a}$}
  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\gias}{$g_\text{ias}$}
  \newcommand{\gmax}{$g_\text{max}$}
  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
  \newcommand{\gsw}{$g_\text{sw}$}
  \newcommand{\loggsw}{$\log(g_\text{sw})$}
  \newcommand{\logA}{$\log(A)$}
  \newcommand{\ppfd}{$\mathrm{PPFD}$}
  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\rh}{$\mathrm{RH}$}
  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
mainfont: Times New Roman
bibliography: solanum-aa.bib
csl: science.csl
---

@@

The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts. Amphistomatous leaves are common among herbaceous plants growing in sunny habitats, including many crop species. This distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand. We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides*). Plants grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity. Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats. Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature. We estimate that hypostomatous leaves would expend 10-65\% more water to achieve the same photosynthetic rate as a comparable amphistomatous leaf, implying large fitness costs of amphistomy in  ecological contexts where this trait is rare.

@@

<!-- Paragraph 1 -->
Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells. They are essential for balancing carbon gained per unit water lost and were an essential innovation in vascular plants, permitting them to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022]. Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020]. Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation [@woodward_stomatal_1987; @buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], predicting paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], and improving crops [@hofmann_impact_2025]. Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].

@@

<!-- Paragraph 3 -->
The paradoxical fact is that, despite the photosynthetic benefit, most leaves are not amphistomatous. Many vertically oriented and/or isobilateral leaves are amphistomatous [@wood_physiology_1934; @howell_concerning_1945, @drake_two_2019]. But among dorsiventral leaves, it is primarily herbaceous plants in open, high light habitats that tend to have amphistomatous leaves [@salisbury_i_1928; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025]. Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface. Even resupinate leaves develop stomata on the lower, albeit adaxial surface [@lyshede_comparative_2002], suggesting that leaf orientation (lower vs. upper) rather than leaf polarity (abaxial vs. adaxial) is causal. The covariation between stomatal density ratio and light habitat is both qualitative and quantitative. A higher proportion of sun leaves are amphistomatous [@salisbury_i_1928] and the proportion of stomata on the upper, adaxial surface increases with light [@bucher_stomatal_2017; @muir_light_2018]. Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.

@@

```{r, concept, echo = FALSE}
#| label: fig-concept
#| fig-align: center
#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatory hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI). The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI. The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."

@@

```{r, aa_cap, echo = FALSE, warning=FALSE, message=FALSE} 
#| label: fig-aacap
#| fig-align: center
#| fig-width: 0.1
#| fig-height: 0.1
#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) in greater wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf. In both panels, estimates are shown for plants grown under simulated shade (brown) or sun (orange) and measured under high light intensity (\\ppfdequals{2000}; open circles) and low light intensity (\\ppfdequals{150}; solid circles). The dashed line at zero indicates no difference in photosynthesis between amphi- and pseudohypostomatous leaves. (\\textbf{A}) The points are the posterior median \\aax{} for each population, with error bars showing 95% confidence intervals. Within each facet, the populations are arranged by \\aax{} estimated in that growth and measurement condition. The average \\aax{} among populations for each treatment group is written in the top left corner. (\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity ploted against native plant area index (PAI; $x$-axis, log-scale). The confidence intervals were omitted for visual clarity. The median relationship between native PAI ($x$-axis; log-scale) and \\aax{} is shown as a solid line with the shaded region between dashed lines showing the 95% confidence ribbon. The * in the top left corner of each facet indicates the slope of the linear regression is significantly different from zero; n.s. indicates not significant."

@@

<!-- Paragraph 7 (summarize secondary results) -->
The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, contradicts the acclimation hypothesis, and provides modest support for the constitutive hypothesis. Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}. LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019]. While LMA is weakly, albeit significantly (@tbl-fit_aa2), associated with individual-level \aax{} ([Fig. @fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison (@tbl-aa_loo1). However, the direct effect of the sun treatment on \aax{} was weaker when LMA was included in the model (@tbl-fit_aa1 vs. @tbl-fit_aa2), suggestive of a mediating role. Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}. The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972]. At low light intensity, the bulk properties of the leaf are weighed toward the upper palisade where most light is intercepted. Hence, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover. Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface. Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.

@@

We conclude that both developmental plasticity and adaptation to open habitats, but not acclimation, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives. This result changes our understanding by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather arise as a byproduct of anatomical and biochemical changes caused by plasticity and adaptation to higher light intensity modulate \aax{}. To gain more precise understanding of when amphistomy is most beneficial will require further research on its leaf anataomical and biochemical basis.

The magnitude of \aax{} we document across populations is also noteworthy, as it implies that hypostomatous leaves predominant in mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss. The cost of hypostomy can be quantified as the difference in total \gsw{} that an amphistomatous leaf would require to achieve the same photosynthetic rate. This cost can be locally approximated on a log-ratio scale as $\mathrm{AA} / \varepsilon_g$ (Supporting Information), where $\varepsilon_g$ is the elasticity of $A$ to \gsw. For $\mathrm{AA} = 0.05$, the water cost of hypostomy would be 0.5 (64.9\%) at low elasticities ($\varepsilon_g = 0.1$) and 0.1 (10.5\%) at high elasticities ($\varepsilon_g = 0.5$). For plants to expend this much extra water implies a large fitness cost of upper stomata in certain ecological contexts. As this is the first quantitative comparison of \aax{} across species, estimation from a broader range of species and environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.

@@

We compared \aax{} among `r n_acc` ecologically diverse populations of wild tomato, including representatives of all described species of *Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides* [@peralta_taxonomy_2008] and the cultivated tomato *S. lycopersicum* var. *lycopersicum* (@tbl-populations). Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`. Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession. [anything else to say here? maybe explain population selection and phylogeny?]

@@

In all growth spaces, we recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All environmental measurements were taken every 10 minutes from the middle of plants racks at approximately the same height as the leaves we measured. We measured leaf temperature of focal leaves prior to measurement using an infrared radiometer (SI-111-SS, Apogee Instruments, Logan, Utah, USA).

@@

We assigned sun and shade treatment to lower and upper racks of a $\qty{1.22}{\meter} \times \qty{2.44}{\meter}$ shelving unit in a climate-controlled growth room. We assigned the sun treatment to the lower rack to limit diffuse light from reaching the shade treatment. The average daytime \ppfd{} was \ppfdqty{`r ppfd_sun`} and \ppfdqty{`r ppfd_shade`} for sun and shade treatments, respectively. To isolate the effect of light intensity from quality, we used the same LED model with the the same spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the shade treatment. To maintain homogeneous environmental conditions other than light, we mixed air within the growth room using an air circulator (Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA). Despite these efforts, the air in the sun treatment was on average $\qty{`r temp_diff`}{\degreeCelsius}$ warmer and the average RH was consequently `r rh_diff`% lower. However, because of evaporative cooling, the leaves in the sun treatment were only $\qty{`r leaf_temp_diff`}{\degreeCelsius}$ on average ($n = `r n_leaf_temp`$ leaves).

@@

All models were using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than `r aa_args[["max_rhat"]]` and the bulk effective sample size (ESS) to greater than `r aa_args[["min_ess"]]` for all model parameters, and there were fewer than `r aa_args[["max_divergent"]]` divergent transitions.

>>>>

---
title: Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves
format:
  pdf:
    keep-tex: true  
    toc: false
author:
  - name: "Christopher D. Muir$^{1,2*}$"
    orcid: 0000-0003-2555-3878
  - name: "Wei Shen Lim$^{1}$"
  - name: "Dachuan Wang$^{1,2}$"
affiliations: false
number-sections: false
header-includes: |
  \usepackage{hyperref}
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
  \newcommand{\aalow}{\mathrm{AA}_{150}}
  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
  \newcommand{\aax}{$\mathrm{AA}$}
  \newcommand{\aay}{\mathrm{AA}}
  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
  \newcommand{\aqcurve}{$A \textendash Q$}
  \newcommand{\ca}{$C_\mathrm{a}$}
  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\gias}{$g_\text{ias}$}
  \newcommand{\gmax}{$g_\text{max}$}
  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
  \newcommand{\gsw}{$g_\text{sw}$}
  \newcommand{\loggsw}{$\log(g_\text{sw})$}
  \newcommand{\logA}{$\log(A)$}
  \newcommand{\ppfd}{$\mathrm{PPFD}$}
  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\rh}{$\mathrm{RH}$}
  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
mainfont: Times New Roman
bibliography: solanum-aa.bib
csl: science.csl
---

@@

<!-- Abstracts of Research Articles should explain to the general reader why the research was done, what was found and why the results are important. They should start with some brief BACKGROUND information: a sentence giving a broad introduction to the field comprehensible to the general reader, and then a sentence of more detailed background specific to your study. This should be followed by an explanation of the OBJECTIVES/METHODS and then the RESULTS. The final sentence should outline the main CONCLUSIONS of the study, in terms that will be comprehensible to all readers. The Abstract is distinct from the main body of the text and thus should not be the only source of background information critical to understanding the manuscript. Please do not include citations or abbreviations in the Abstract. The abstract should be 125 words or less. For Perspectives and Policy Forums, please include a one-sentence abstract. -->

The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts. Paradoxically, despite this photosynthetic advantage, most leaves are hypostomatous (stomata on lower surface only). Based on the observation that amphistomatous leaves are most common among herbaceous plants growing in sunny habitats, including many crop species, it has long been hypothesized that traits affecting CO$_2$ supply or demand amplify the advantage of amphistomy under high light intensity. We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides*). Plants grown under high light intensity benefit more from amphistomy in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity. Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats. Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature. We estimate that hypostomatous leaves would expend 10-65\% more water to achieve the same photosynthetic rate as a comparable amphistomatous leaf, implying large fitness costs of amphistomy in ecological contexts where this trait is rare.

@@

<!-- Paragraph 1 -->
Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells. They are essential for balancing carbon gained per unit water lost and were an essential innovation in vascular plants, permitting them to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022]. Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020]. Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for studying adaptation [@buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013], inferring paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], predicting responses to climate change [@woodward_stomatal_1987; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], and improving crops [@hofmann_impact_2025]. Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].

@@

<!-- Paragraph 3 -->
The paradoxical fact is that, despite the photosynthetic benefit, most leaves are not amphistomatous. Many vertically oriented and/or isobilateral leaves are amphistomatous [@wood_physiology_1934; @howell_concerning_1945, @drake_two_2019]. But among dorsiventral leaves, it is primarily herbaceous plants in open, high light habitats that tend to have amphistomatous leaves [@salisbury_i_1928; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025]. Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface. Even resupinate leaves develop stomata on the lower, albeit adaxial surface [@lyshede_comparative_2002], suggesting that leaf orientation (lower vs. upper) rather than leaf polarity (abaxial vs. adaxial) is causal. Stomatal density ratio, defined as the ratio of upper to total stomatal density, is a quantitative metric of stomatal distribution. The covariation between stomatal density ratio and light habitat is both qualitative and quantitative. A higher proportion of sun leaves are amphistomatous [@salisbury_i_1928] and the proportion of stomata on the upper, adaxial surface increases with light [@bucher_stomatal_2017; @muir_light_2018]. Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.

@@

```{r, concept, echo = FALSE}
#| label: fig-concept
#| fig-align: center
#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatopn hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI). The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI. The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."

@@

```{r, aa_cap, echo = FALSE, warning=FALSE, message=FALSE} 
#| label: fig-aacap
#| fig-align: center
#| fig-width: 0.1
#| fig-height: 0.1
#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) is greater in wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf. In both panels, estimates are shown for plants grown under simulated shade (brown) or sun (orange) and measured under high light intensity (\\ppfdequals{2000}; open circles) and low light intensity (\\ppfdequals{150}; solid circles). The dashed line at zero indicates no difference in photosynthesis between amphi- and pseudohypostomatous leaves. (\\textbf{A}) The points are the posterior median \\aax{} for each population, with error bars showing 95% confidence intervals. Within each facet, the populations are arranged by \\aax{} estimated in that growth and measurement condition. The average \\aax{} among populations for each treatment group is written in the top left corner. (\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity plotted against native plant area index (PAI; $x$-axis, log-scale). The confidence intervals were omitted for visual clarity. The median relationship between native PAI ($x$-axis; log-scale) and \\aax{} is shown as a solid line with the shaded region between dashed lines showing the 95% confidence ribbon. The * in the top left corner of each facet indicates the slope of the linear regression is significantly different from zero; n.s. indicates not significant."

@@

<!-- Paragraph 7 (summarize secondary results) -->
The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, contradicts the acclimation hypothesis, and provides modest support for the constitutive hypothesis. Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}. LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019]. While LMA is weakly, albeit significantly (@tbl-fit_aa2), associated with individual-level \aax{} ([Fig. @fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison (@tbl-aa_loo1). However, the direct effect of the sun treatment on \aax{} was weaker when LMA was included in the model (@tbl-fit_aa1 vs. @tbl-fit_aa2), suggestive of a mediating role. Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed to  identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}. The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972]. At low light intensity, photosynthesis is weighted toward the upper palisade where most light is intercepted. If resistance to CO$_2$ diffusion is high in the upper palisade, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover. Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface. Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.

@@

We conclude that both developmental plasticity and adaptation to open habitats, but not acclimation, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives. This result changes our understanding by showing that high light intensity *per se* does not increase the benefit of amphistomy. Instead, the benefit increases as a byproduct of anatomical and biochemical changes caused by plasticity and adaptation to higher light intensity. To gain more precise understanding of when amphistomy is most beneficial will require further research on its leaf anatomical and biochemical basis.

The magnitude of \aax{} we document across populations is also noteworthy, as it implies that hypostomatous leaves predominant in mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss. The cost of hypostomy can be quantified as the difference in total \gsw{} that an amphistomatous leaf would require to achieve the same photosynthetic rate. This cost can be locally approximated on a log-ratio scale as $\mathrm{AA} / \varepsilon_g$ (Supporting Information), where $\varepsilon_g$ is the elasticity of $A$ to \gsw. For $\mathrm{AA} = 0.05$, the water cost of hypostomy would be 0.5 (64.9\%) at low elasticities ($\varepsilon_g = 0.1$) and 0.1 (10.5\%) at high elasticities ($\varepsilon_g = 0.5$). For plants to expend this much extra water implies a large fitness cost of upper stomata in certain ecological contexts. As this is the first quantitative comparison of \aax{} across species, estimates from a broader range of species and more stressful environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.

@@

We compared \aax{} among `r n_acc` ecologically diverse populations of wild tomato, including representatives of all described species of *Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides* [@peralta_taxonomy_2008] and the cultivated tomato *S. lycopersicum* var. *lycopersicum* (@tbl-populations). These populations were selected in a previous study because they encompass the breadth of climatic variation in the wild tomato clade [@pease_phylogenomics_2016]. Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`. Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.

@@

We recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All environmental measurements were taken every 10 minutes from the middle of plants racks at approximately the same height as the leaves we measured. We measured leaf temperature of focal leaves prior to measurement using an infrared radiometer (SI-111-SS, Apogee Instruments, Logan, Utah, USA).

@@

We assigned sun and shade treatment to lower and upper racks of a $\qty{1.22}{\meter} \times \qty{2.44}{\meter}$ shelving unit in a climate-controlled growth room. We assigned the sun treatment to the lower rack to limit diffuse light from reaching the shade treatment. The average daytime \ppfd{} was \ppfdqty{`r ppfd_sun`} and \ppfdqty{`r ppfd_shade`} for sun and shade treatments, respectively. To isolate the effect of light intensity from quality, we used the same LED model with the the same spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the shade treatment. To maintain homogeneous environmental conditions other than light, we mixed air within the growth room using an air circulator (Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA). Despite these efforts, the air in the sun treatment was on average $\qty{`r temp_diff`}{\degreeCelsius}$ warmer and the average RH was consequently `r rh_diff`% lower. However, because of evaporative cooling, the leaves in the sun treatment were only $\qty{`r leaf_temp_diff`}{\degreeCelsius}$ warmer on average ($n = `r n_leaf_temp`$ leaves).

@@

All models were fit using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than `r aa_args[["max_rhat"]]` and the bulk effective sample size (ESS) to greater than `r aa_args[["min_ess"]]` for all model parameters, and there were fewer than `r aa_args[["max_divergent"]]` divergent transitions.
